NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-02-08 Manuscript/manuscriptPS.tex
Update manuscriptPS.tex

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eukaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of red blood cells,
but the abundance varies between different organelles up to
25-35\% in the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance \todo{Relatively low compared to what?}, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains interact specifically with PS lipids,
while others are attracted by nonspecific electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of cation binding
are crucial for the PS-mediated signaling processes.

Previous experimental studies have concluded that the
PS headgroup is more rigid than the phosphatidylcholine (PC)
due to hydrogen bonding network or
electrostatic interactions \cite{browning80,buldt81}.
Multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids,
while monovalent ions interact more weakly with
PS-containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
The dehydrated complexes of PS headgroup and calcium ions can also lead to
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
On the other hand, some studies propose that the specific binding
affinity is similar for the negatively charged and zwitterionic lipids, and
the increased cation binding to negatively charged lipid bilayers arises only due
to the increased local cation concentration in the membrane vicinity~\cite{seelig90,sinn06}.
Dilution of bilayers with PC lipids makes PS headgroups
less rigid and reduces their propensity to form
strong complexes with multivalent ions~\cite{browning80,buldt81,roux90,roux91}.
The molecular level interpretation of these observations is,
however, lacking.

@@

Several classical molecular dynamics (MD) simulation studies have been done
to understand the PS headgroups, their influence on lipid bilayer properties, and their
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
However, the results have depended strongly on the force field used.
For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
the same force field without the NBfix parameters, combined with NMR chemical shifts and
REDOR \todo{Spell out REDOR?} experiments, suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.
We have recently demonstrated that such controversies can be resolved by
comparing the C--H bond order parameters, $S_\mathrm{CH}$, of lipid headgroups between simulations
and experiments \cite{botan15,catte16}. The $S_\mathrm{CH}$ can be directly
measured from NMR experiments with high accuracy and compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}.
Previous studies showed that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}.
%
Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
\todo{Should we leave the mention of ECC out from the Introduction
(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

Here, we collect a set of experimentally measured lipid headgroup and
glycerol backbone C--H bond order parameters, which can be used to
evaluate the quality of headgroup structure and ion binding affinity
in MD simulations of lipid bilayers containing PS lipids. 
The available MD simulation models of PS are then compared with
the collected experimental data.
%using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
The results pave the way
for development of lipid force fields with realistic description of
the headgroup region of negatively charged lipids in physiological salt
conditions. Such models are expected to be useful in understanding
biological function of lipid headgroups and glycerol backbone, as
these behave similarly in model membranes and in bacterial cells \cite{gally81,scherer87,seelig90}.

>>>>

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eukaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of red blood cells. 
The abundance, however, varies between different organelles, and up to
25-35\% of the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14} consists of PS lipids.
PS lipids are vastly important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains interact specifically with PS lipids,
while others are attracted by nonspecific electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of cation binding
are crucial for the PS-mediated signaling processes.

Experimental studies have indicated that the
PS headgroup is more rigid than the phosphatidylcholine (PC)
due to hydrogen bonding network or
electrostatic interactions \cite{browning80,buldt81}.
While monovalent ions interact weakly with
PS-containing bilayers, multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}. The dehydrated complexes of PS headgroup and calcium ions can even lead to
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
On the other hand, some studies propose that the specific binding
affinity is similar for the negatively charged and zwitterionic lipids, and
the increased cation binding to negatively charged lipid bilayers arises only due
to the increased local cation concentration in the membrane vicinity~\cite{seelig90,sinn06}.
Dilution of bilayers with PC lipids makes PS headgroups
less rigid and reduces their propensity to form
strong complexes with multivalent ions~\cite{browning80,buldt81,roux90,roux91}.

@@

The molecular level interpretation of these observations is,
however, lacking, and  classical molecular dynamics (MD) simulations have been widely used in efforts to
to understand the PS headgroup structure, their influence on lipid bilayer properties, and their
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
Unfortunately, the results have depended strongly on the force field used.
For example, recent simulations using the NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
the same force field without the NBfix parameters, combined with NMR chemical shifts and
REDOR \todo{Spell out REDOR?} experiments, suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

We have recently demonstrated that such controversies can be resolved by
comparing the C--H bond order parameters, $S_\mathrm{CH}$, of lipid headgroups between simulations
and experiments \cite{botan15,catte16}. The $S_\mathrm{CH}$ can be directly
measured from NMR experiments with high accuracy and compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}.
Previous studies showed that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}.
%
Based on these data, the cation binding affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability using the electronic continuum correction \cite{melcr18}.
\todo{Should we leave the mention of ECC out from the Introduction
(i.e., mention it only in the Conclusions) as the ECC parameters are not used in the paper?}

Here, we collect a set of experimentally measured lipid headgroup and
glycerol backbone C--H bond order parameters, which can be used to
evaluate the quality of headgroup structure and ion binding affinity
in MD simulations of lipid bilayers containing PS lipids. 
The available MD simulation models of PS are then compared with
the collected experimental data.
%using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
The results pave the way
for development of lipid models that correctly describe 
the headgroup region of negatively charged lipids in physiological salt
conditions. Such force fields are expected to be useful in understanding
biological function of lipid headgroups and glycerol backbone, as
these are knwon to behave similarly in simple model membranes and in cells \cite{gally81,scherer87,seelig90}.
